A Comparison of Neural Network Models for Water Quality Forecasting

### 2024-06-28 manuscript/manuscript.Rmd
End of day edits

This study had the simple motivation of comducting a model comparison, and has resulted in an abundance of research questions and ideas for future work. Outside of exploring how the performance of neural networks will vary with forecast horizon, other promising directions for exploration include cross-learning and hybrid models. Throughout this study, we trained individual models for each target time series. Cross-learning presents a different approach where we would train one model across multiple time series (potentially, all the time series for a give target variable or even across target variables). Through cross-learning, the model might be able to learn patterns from one time series that will transfer to making accurate forecasts on a different time series. Along these lines, developing a model hybridized across modeling classes could yield superior performance. Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model alone. For instance, one could explore a model that heavily relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the  when  longer horizons. This motivates the exploration of a hybrid model that employs AutoTheta at the beginning of the forecast horizon and neural network models later in the forecast horizon.

It is an exciting time to study water quality forecasting. There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as the cost for sensors and remote sensing continue to come down. Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns from data without being restrained by a need for domain knowledge. Furthermore, their performance has the potential to improve as more water quality data becomes available. Yet, they also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management. This study does not attempt to provide a definitive answer as to whether neural networks will be the best method class moving forward as answering this question will take a plethora of research. Instead, we hope that this study will generate more ideas and incite momentum towards improving our ability to forecast and manage water quality.

>>>>

This study had the sole motivation of conducting a model comparison, and has resulted in an abundance of research questions and ideas for future work. Outside of exploring how the performance of neural networks will vary with forecast horizon, other promising directions for exploration include cross-learning and hybrid models. Throughout this study, we trained individual models for each target time series. Cross-learning presents a different approach where one model would be trained across multiple time series. Recently, cross-learning has been shown to improved forecast performance as models that employ cross-learning are able to learn patterns across different time series that transfer to making more accurate forecasts on an individual time series, particularly when there is a limited amount of data points in the individual time series  \cite{time_series}. Along these lines, developing a model hybridized across modeling classes could also yield superior performance. Since we observed that AutoTheta was the best performing model early in the forecast horizon, and neural networks were the best performers later in the forecast horizon, it is sensible that a hybrid ML-statistical model could achieve better performance than a purely statistical or machine learning model. For instance, a model that relies on a statistical model the beginning of a forecast horizon then favors a ML model towards the end of forecast window could achieve better performance than the models explored in this study.

It is an exciting time to study water quality forecasting. There are large, publicly accessible water quality data sets that exist now, and there will be an increasing amount of limnological data in the future as the cost for sensors and remote sensing continue to decline. Neural network models offer promising advantages over alternative methods for limnological forecasting as these methods can learn complex patterns from data without being restrained by a need for domain knowledge. Furthermore, the performance of neural networks has the potential to improve as more water quality data becomes available. Yet, neural networks also present a range of problems like a lack of interpretability and generalizability that could restrain them from being useful decision support tools for the safety critical problem of water resource management. This study does not attempt to provide a definitive answer as to whether neural networks will be the best method class moving forward as resolving such a question will require a much larger body of work than what is presented here. Instead, the hope is that this study will generate ideas and incite momentum towards improving our ability to forecast and manage water quality.
